Seasonality in governing dynamics at submesoscales in the Kuroshio Extension

### 2016-06-28 writeup/rocha_etal.tex
Small edits

\begin{abstract}
    Two new high-resolution numerical simulations with embedded tides suggest a
    strong modulation in the governing near-surface dynamics at submesoscales
    (roughly 10-100 km) in the Kuroshio Extension. Consistent with recent studies, deep late-Winter mixed layers
    are prone to baroclinic instabilities, and submesoscale turbulence
    prevails in late Winter/early Spring. Inertia-gravity waves are enhanced near
    the surface when the upper ocean is strongly stratified. Thus the summertime
    re-stratification weakens submesoscale turbulence and enhances intertia-gravity
     waves near the surface. In the Kuroshio Extension,
    inertia-gravity waves strongly dominate the surface kinetic energy and
    sea-surface height variance in late Summer/early Fall at submesoscales.
\end{abstract}

@@

\section{The LLC numerical simulations}
We use the output of two latitude-longitude polar cap (LLC)
realistic numerical simulations. The outputs
analyzed here, LLC2160 (1/24$^\circ$) and LLC4320 (1/48$^\circ$)  have
90-vertical-levels and are
forward Massachusetts Institute of Technology general circulation model (MITgcm)
numerical solutions on a LLC grid   \citep{forget_etal2015} that were
spun up from a ECCO2 adjoint-method state estimate. Both simulations were forced by
tides and 6-hourly surface fluxes. The LLC2160
output spans two years from March 2011 to April 2013; the LLC4320 was spun up from
the LLC2160 simulation, spanning one year from September 2011 to September 2012.
The LLC4320 simulation is an extension of the 5-month long output used by
\citet{rocha_etal2016}.

@@

To study seasonal variations in the upper-ocean dynamics, we focus on the northwest
Pacific, in the vicinity of the Kuroshio
Extension, where previous studies have suggested strong submesoscale seasonality
\citep{sasaki_etal2014,qiu_etal2014}.
We analyze a sub-domain of the LLC4320 and LLC2160 simulations of about 2000 km$^2$
spanning 155-175$^\circ$E; 25-40$^\circ$N (figure \ref{fig1}a). The stratification
in this mesoscale-rich subtropical region undergoes a vigorous seasonal cycle: wintertime-enhanced
small-scale turbulence de-stratifies the upper ocean, yielding mixed layers
as deep as 300m. In late Spring/early Summer enhanced solar radiation and mixed-layer
instabilities re-stratify the upper ocean, yielding mixed layers as shallow as 40m.
Fundamentally, the stratification and its seasonality are well-captured by both LLC simulations:
a comparison with Argo climatology shows that both simulations skillfully represent the Kuroshio
Extension stratification and its seasonal variability  (supplemental material).

@@

Figures \ref{fig1}b-c show snapshots of vertical vorticity $\zeta$ in early Spring
(April 15) and Fall (October 15) in the LLC4320 (1/48$^\circ$) simulation.
The model solutions depict seasonality in vorticity: large values of
fine-grained vertical vorticity are observed in early Spring with maximum
values as large as $4f$, where $f$ is
the local planetary vorticity, and root-mean-square (RMS) of about $0.4f$. In early
Fall, the situation is the opposite: the vertical vorticity is relatively coarse-grained,
its local maximum and RMS is smaller than $0.5f$.
Indeed, the seasonal cycle of vorticity and rate of strain are strong (figure~\ref{fig2}):
in both simulations, the RMS vorticity and strain are about twice as large in
late Winter/early Spring
than in late Summer/early Fall. Because the wintertime vorticity and strain rate
are dominated by the smallest scales in the flow, increasing the resolution from
1/24$^\circ$ to 1/48$^\circ$ significantly increases the wintertime RMS vorticity
and strain by about 40$\%$.

 \begin{figure*}[ht]
   \begin{center}
     \includegraphics[width=.65\textwidth]{figs/fig2.pdf}
  \caption{Time series of the root-mean-square (RMS) of vorticity (a) and
  lateral rate of strain (b), and horizontal divergence (c) in the LLC outputs and gridded AVISO data.}
  \label{fig2}
  \end{center}
\end{figure*}

The bulk of vertical vorticity and lateral strain rate are associated with subinertial flows
($T_f = 2\pi/f_0\approx 23.5$h, where $f_0$ is the inertial frequency at
the mean latitude): daily-averaging the velocity fields suppresses superinertial
motions and reduces the RMS
vorticity by less than 40$\%$ and the RMS strain by less than 10$\%$; the seasonal
cycle remains strong (see
dashed lines in figures \ref{fig2}a-b). Most of this seasonal cycle is associated
with submesoscale flows: smoothing the velocity fields
with a Hanning filter with cut-off scale of 100 km dramatically reduces the RMS
vorticity and strain. The reduction in variance is about 80$\%$ in winter,
yielding RMS vorticity and strain roughly consistent with the diagnostics from
AVISO gridded geostrophic velocities (compare red lines to black lines in figures \ref{fig2}a-b).
The overall picture that emerges is consistent with recent studies: shallow baroclinic
instabilities energize the submesoscales in late Winter, drawing from the available
potential energy stored in deep mixed layers \citep{sasaki_etal2014,callies_etal2015,callies_etal2016}.

@@

\section{Seasonality in governing submesoscale dynamics}
The overall results of figure \ref{fig2}c suggest that submesoscale surface
 variability stems from different dynamics in Summer than Winter.
 To better understand these differences, we calculate
joint probability distributions (jPDF) of
vorticity-strain and vorticity-Laplacian of sea-surface height (\ref{fig3}).

@@

To better understand the projection of these flows onto different horizontal
scales, we calculate wavenumber spectra of kinetic energy and sea-surface height
(SSH) variance.  Before calculating the spectra, time-mean and spatial linear
trends were removed, and the resulting fields were multiplied by a two-dimensional Hanning ``spectral window''.
The two-dimensional spectra were averaged azimuthly \citep[e.g., ][]{rocha_etal2016} .
We discuss only spectra for the 1/48$^\circ$, which extends the 1/24$^\circ$
towards smaller scales (supplemental material).

@@

Figure \ref{fig4}a depicts the horizontal wavenumber spectra of surface KE in
April and October. At scales larger than 20km, the spectra based on hourly
velocity snapshots are nearly indistinguishable within 95$\%$ confidence level
(compare solid lines in figure \ref{fig4}a). Consistent with the results of
\citet{rocha_etal2016} who analyzed a 5-month output of the LLC4320 simulation in Drake Passage,
there is significant high-frequency variabilty at submesoscales. Daily-averaging
the velocity field suppresses spatial variability at scales smaller than about 250
km both in April and October (compare solid lines against dashed lines in figure
\ref{fig4}a). But this suppression is dramatic in October, when the horizontally
divergent flows peak. At scales
smaller than 100 km, 39$\%$ of the surface KE in April is accounted for by super-inertial
flows as opposed to 79$\%$ in October. The seasonality of subinertial submesoscale flows
is more dramatic than the seasonal cycle of the total flow. This is consistent
with the results of \citet{sasaki_etal2014}, which are
based on daily-averaged velocity fields of a different model without tidal forcing
(P. Klein, personal communication).

These high-frequency submesoscale flows significantly project on the sea-surface.
In October, when the inertia-gravity waves are strongest the surface and submesoscale
turbulence is weakest. There is a dramatic difference between the spectra
based on hourly and daily-averaged SSH: at scales smaller than 100 km, the spectra
of hourly SSH roughly follows a $-2$ power-law, whereas the spectra of daily-averaged
SSH roughly follows a $-5$ power-law. At these scales, 33$\%$ of the SSH variance in April is accounted for by suprainertial
flows as opposed to 83$\%$ in October! Curiously, the out-of-phase seasonal cycle of submesoscale
turbulence and near-surface inertia-gravity waves conspire to yield very weak seasonality
in the spectra of KE and SSH variance based on hourly fields.

@@

Our work adds to recent studies that presented modeling \cite{sasaki_etal2014}
and observational \citep{callies_etal2015,buckingham_etal2016} evidence of vigorous seasonality in
submesoscale turbulence. We conjecture that the summertime dominance of inertia-gravity waves
 \citep{callies_etal2015} is a consequence both of suppression of submesoscale turbulence and
enhancement of inertia-gravity waves due to re-stratification of
the upper ocean. Of course, our study focuses on a single patch of ocean in the
vicinity of the Kuroshio Extension, which may be typical of mesoscale-rich
subtropical regions, but not
representative of other regions such as low eddy kinetic energy eastern boundary currents
and the middle of the subtropical gyre; care must be taken to avoid overgeneralizing
these results.  We plan to report on the geographic variability of submesoscale
 seasonality in a future study.

>>>>

\begin{abstract}
    Two new high-resolution numerical simulations with embedded tides suggest a
    strong modulation in the governing near-surface dynamics at submesoscales
    (roughly 10-100 km) in the Kuroshio Extension. Consistent with recent studies, deep late-Winter mixed layers
    are prone to baroclinic instabilities, and submesoscale turbulence
    prevails in late Winter/early Spring. Inertia-gravity waves are enhanced near
    the surface when the upper ocean is strongly stratified. Thus the summertime
    re-stratification weakens submesoscale turbulence and enhances intertia-gravity
     waves near the surface. In the Kuroshio Extension,
    inertia-gravity waves strongly dominate the submesoscale surface kinetic energy and
    sea-surface height variance in late Summer/early Fall.
\end{abstract}

@@

\section{The LLC numerical simulations}
We use the output of two latitude-longitude polar cap (LLC)
realistic numerical simulations. The outputs
analyzed here, LLC2160 (1/24$^\circ$) and LLC4320 (1/48$^\circ$),   are
forward Massachusetts Institute of Technology general circulation model (MITgcm)
numerical solutions on a LLC grid  \citep{forget_etal2015} with
90-vertical-levels. The coarser-resolution LLC simulation was
spun up from a ECCO2 adjoint-method state estimate, which synthetized millions
of observations from 2009 through 2011. Both simulations were forced by
tides and 6-hourly surface fluxes. The LLC2160
output spans two years from March 2011 to April 2013; the LLC4320 was spun up from
the LLC2160 simulation, spanning one year from September 2011 to September 2012.
The LLC4320 simulation is an extension of the 5-month long output used by
\citet{rocha_etal2016}.

@@

To study seasonal variations in the upper-ocean dynamics, we focus on the northwest
Pacific, in the vicinity of the Kuroshio
Extension, where previous studies have suggested strong submesoscale seasonality
\citep{sasaki_etal2014,qiu_etal2014}.
We analyze a sub-domain of the LLC4320 and LLC2160 simulations of about 2000 km$^2$
spanning 155-175$^\circ$E; 25-40$^\circ$N (figure \ref{fig1}a). The stratification
in this mesoscale-rich subtropical region undergoes a vigorous seasonal cycle: wintertime-enhanced
small-scale turbulence de-stratifies the upper ocean, yielding mixed layers
as deep as 300m. In late Spring/early Summer enhanced solar radiation and mixed-layer
instabilities re-stratify the upper ocean, yielding mixed layers as shallow as 40m.
Fundamentally, the upper-ocean density structure is well-captured by both LLC simulations:
a comparison with Argo climatology shows that both simulations skillfully represent the Kuroshio
Extension stratification and its seasonal variability  (supplemental material).

@@

Figures \ref{fig1}b-c show snapshots of vertical vorticity $\zeta$ in early Spring
(April 15) and Fall (October 15) in the LLC4320 (1/48$^\circ$) simulation.
The model solutions depict seasonality in vorticity: large values of
fine-grained vertical vorticity are observed in early Spring with maximum
values as large as $4f$, where $f$ is
the local planetary vorticity, and root-mean-square (RMS) of about $0.4f$. In early
Fall, the situation is the opposite: the vertical vorticity is relatively coarse-grained,
its local maximum and RMS is smaller than $0.5f$.
Indeed, the seasonal cycle of vorticity and rate of strain are strong (figure~\ref{fig2}):
in both simulations, the RMS vorticity and strain are about twice as large in
late Winter/early Spring
than in late Summer/early Fall. Because the wintertime vorticity and strain rate
are dominated by the smallest scales in the flow (the KE spectra is shallower than
a $-3$ power-law in Winter), increasing the resolution from
1/24$^\circ$ to 1/48$^\circ$ significantly increases the wintertime RMS vorticity
and strain by about 40$\%$.

 \begin{figure*}[ht]
   \begin{center}
     \includegraphics[width=.65\textwidth]{figs/fig2.pdf}
  \caption{Time series of the root-mean-square (RMS) of vorticity (a) and
  rate of strain (b), and horizontal divergence (c) in the LLC outputs and gridded AVISO data.}
  \label{fig2}
  \end{center}
\end{figure*}

The bulk of vertical vorticity and lateral strain rate are associated with subinertial flows
($T_f = 2\pi/f_0\approx 23.5$h, where $f_0$ is the inertial frequency at
the mean latitude): daily-averaging the velocity fields suppresses superinertial
motions and reduces the RMS
vorticity by less than 40$\%$ and the RMS strain by less than 10$\%$; the seasonal
cycle remains strong (see
dashed lines in figures \ref{fig2}a-b). Most of this seasonal cycle is associated
with submesoscale flows: smoothing the velocity fields
with a Hanning filter with cut-off scale of 100 km dramatically reduces the RMS
vorticity and strain. The reduction in variance is about 80$\%$ in winter,
yielding RMS vorticity and strain roughly consistent with the diagnostics from
AVISO gridded geostrophic velocities (compare red lines to black lines in figures \ref{fig2}a-b).
The picture that emerges is consistent with recent studies: shallow baroclinic
instabilities energize the submesoscales in late Winter, drawing from the available
potential energy stored in deep mixed layers \citep{sasaki_etal2014,callies_etal2015,callies_etal2016}.

@@

\section{Seasonality in governing submesoscale dynamics}
The overall results of figure \ref{fig2}c suggest that submesoscale surface
 variability stems from different dynamics in Summer than Winter.
 To characterize these differences, we calculate
joint probability distributions (jPDF) of
vorticity-strain and vorticity-Laplacian of sea-surface height (\ref{fig3}).

@@

To better quantify the projection of these flows onto different horizontal
scales, we calculate wavenumber spectra of kinetic energy and sea-surface height
(SSH) variance.  Before calculating the spectra, time-mean and spatial linear
trends were removed, and the resulting fields were multiplied by a two-dimensional Hanning ``spectral window''.
The two-dimensional spectra were averaged azimuthly \citep[e.g., ][]{rocha_etal2016} .
We discuss only spectra for the 1/48$^\circ$, which extends the 1/24$^\circ$
towards smaller scales (supplemental material).

@@

Figure \ref{fig4}a depicts the horizontal wavenumber spectra of surface KE in
April and October. At scales larger than 20km, the spectra based on hourly
velocity snapshots are nearly indistinguishable within 95$\%$ confidence level
(compare solid lines in figure \ref{fig4}a). Consistent with the results of
\citet{rocha_etal2016} who analyzed a 5-month output of the LLC4320 simulation in Drake Passage,
there is significant high-frequency variabilty at submesoscales. Daily-averaging
the velocity field suppresses spatial variability at scales smaller than about 250
km both in April and October (compare solid lines against dashed lines in figure
\ref{fig4}a). But this suppression is dramatic in October, when the inertia-gravity waves peak. At scales
smaller than 100 km, 39$\%$ of the surface KE in April is accounted for by super-inertial
flows as opposed to 79$\%$ in October. The seasonality of subinertial submesoscale flows
is more dramatic than the seasonal cycle of the total flow. This is consistent
with the results of \citet{sasaki_etal2014}, which are
based on daily-averaged velocity fields of a different model without tidal forcing
(P. Klein, personal communication).

These high-frequency submesoscale flows significantly project on the sea-surface.
In October, when the inertia-gravity waves are strongest the surface and submesoscale
turbulence is weakest. There is a dramatic difference between the spectra
based on hourly and daily-averaged SSH: at scales smaller than 100 km, the spectra
of hourly SSH roughly follows a $-2$ power-law, whereas the spectra of daily-averaged
SSH roughly follows a $-5$ power-law. At these scales, 33$\%$ of the SSH variance in April is accounted for by super-inertial
flows as opposed to 83$\%$ in October! Curiously, the out-of-phase seasonal cycle of submesoscale
turbulence and near-surface inertia-gravity waves conspire to yield very weak seasonality
in the spectra of KE and SSH variance based on hourly fields.

@@

Our work adds to recent studies that presented modeling \cite{sasaki_etal2014}
and observational \citep{callies_etal2015,buckingham_etal2016} evidence of vigorous seasonality in
submesoscale turbulence. We conjecture that the summertime dominance of inertia-gravity waves
 \citep{callies_etal2015} is a consequence both of suppression of submesoscale turbulence and
enhancement of inertia-gravity waves due to re-stratification of
the upper ocean. Of course, our study focuses on a single patch of ocean in the
vicinity of the Kuroshio Extension, which may be typical of mesoscale-rich
subtropical regions, but is unlikely to be 
representative of other regions such as low eddy kinetic energy eastern boundary currents
and the middle of the subtropical gyre; care must be taken to avoid overgeneralizing
these results.  We plan to report on the geographic variability of submesoscale
 seasonality in a future study.
